### 2023-02-27 code/analysis_code/2_sex.Rmd
comments

```{r}
library(here)
library(tidyverse)
library(spdep)
```
https://github.com/ubikuity/List-of-neighboring-states-for-each-US-state
List-of-neighboring-states-for-each-US-state
```{r}
sex_df <- readRDS(here("data/processed_data/sex_new.rds"))
us_nabs <- read_csv(here("data/raw_data/neighbors-states.csv"))

>>>>

```{r}
library(here)
library(tidyverse)
library(spdep)
```
https://github.com/ubikuity/List-of-neighboring-states-for-each-US-state
List-of-neighboring-states-for-each-US-state
```{r}
sex_df <- readRDS(here("data/processed_data/sex_new.rds"))
us_nabs <- readRDS(here("data/processed_data/us_nabs.rds"))

### 2023-02-27 code/processing_code/1_cleaning_2020.Rmd
eda
comments

```{r, echo=FALSE, message=FALSE}
library(knitr)
library(readxl)
library(lubridate)
library(stringr)
library(tidyverse)
library(here)
```

```{r}
deaths <- read_xlsx(here("data/raw_data/AIDSVu_State_Death_2020.xlsx"), skip = 3)
us_race_pop <- read_xlsx(here("data/raw_data/DECENNIALPL2020.P2-2023-02-24T124235.xlsx"), sheet = 2)
us_sex_pop <-read_csv(here("data/raw_data/raw_data.csv"), skip = 2)
kid_pop <- read_xlsx(here("data/raw_data/Child population by age group.xlsx"))
kid_race <- read_xlsx(here("data/raw_data/Child population by race and ethnicity.xlsx"))

@@

Let's start by separating `deaths` by demographic characteristic . While this data provides information on demographic characteristics such as sex, age group, and race/ethnicity, it does not provide data on all of these terms together. We'll separating by sex and race/ethnicity. This analysis will be limited to Non-Hispanic Black, Non-Hispanic White, and Hispanic persons.
```{r}
deaths <- deaths %>% 
  filter(!grepl("DC", `State Abbreviation`)) %>%
  filter(!grepl("PR", `State Abbreviation`))

@@

summary(deaths_race)
```
```{r}
new_dx <- new_dx %>% 
  filter(!grepl("DC", `State Abbreviation`)) %>%
  filter(!grepl("PR", `State Abbreviation`)) %>% 
  select(1:3, contains(c("male", "Black", "White", "Hispanic", "Multiple")), -contains(c("and", "Stability")))

@@

Some rates are -9 due to missingness. Let's convert these to NA
```{r}
deaths_race <- deaths_race %>% 
  mutate(across(ends_with("Death Rate"), ~ case_when( .x < 0 ~ NA, TRUE ~ .x)))
```

Next, let's move on to the census data. I'll start by re-formatting the data from wide format for States to long, and format race/ethnicity into wide.
```{r}
glimpse(us_race_pop)

@@

```{r}
race_pop <- race_pop %>% 
  select(c(Location, starts_with(c("Total:", "Black or African American alone", "Hispanic", "White alone", "Population of two or more races:")))) %>% 
  rename(c(Hispanic = `Hispanic or Latino`,
         White = `White alone`,
         Black = `Black or African American alone`,
         Multiple = `Population of two or more races:`))
```

@@

Then, I'll calculate the counts of males and females in each location.
```{r}
pop_sex <- inner_join(total_pop, us_sex_pop)
```

Next, we need to estimate the population older than 12. Since I cannot find a source explicitly population older than 12, I will estimate population younger than 13 this using data from Kids Count. 
```{r}
kid_pop <- kid_pop %>% 
  filter(TimeFrame == 2020) %>% 
  filter(`Age group` %in% c("0 to 4", "5 to 11", "12 to 14")) %>% 
  mutate(Data = as.numeric(Data))

@@

```{r}
new_pop <- full_join(pop_sex, pop_less13, by = join_by(Location))

@@

```{r}
sex_df <- full_join(deaths_sex, df_pop, by = join_by(State == Location))

sex_df <- full_join(sex_df, sex_dx)
```

```{r}
glimpse(kid_race)

@@

Let's estimate the number of children under 13 by race
```{r}
kid_race <- kid_race %>%
  mutate(across(2:5, ~ round( . * Less13)))

@@

```{r}
race_df <- list(deaths_race, late_dx, new_dx)

@@

```{r}
rec_care <- rec_care %>%
  filter(!grepl("DC", `State Abbreviation`)) %>%
  filter(!grepl("PR", `State Abbreviation`)) %>% 
  select(1:3, contains("Percent of")) %>% 
  select(1:3, contains(c("Hispanic", "Black", "White", "Multiple","male")))

sex_df$`Male_rec_care_percent`<- rec_care$`Percent of Males Living with HIV who Received HIV Care`
sex_df$`Female_rec_care_percent`<- rec_care$`Percent of Females Living with HIV who Received HIV Care`

@@

race_df <- inner_join(race_df, rec_care)
```
```{r}
v_sup <- v_sup %>%
  filter(!grepl("DC", `State Abbreviation`)) %>%
  filter(!grepl("PR", `State Abbreviation`)) %>% 
  select(1:3, contains("Percent of")) %>% 
  select(1:3, contains(c("Hispanic", "Black", "White", "Multiple","male")))

@@

```{r}
prev <- prev %>% 
  filter(!grepl("DC", `State Abbreviation`)) %>%
  filter(!grepl("PR", `State Abbreviation`))

@@

```{r}
saveRDS(sex_df, here("data/processed_data/sex_df.rds"))
saveRDS(race_df, here("data/processed_data/race_df.rds"))
```

```{r}
rm(deaths, df_pop, kid_pop, pop_sex, pop_less13, us_sex_pop, us_race_pop, new_pop, deaths_sex, race_pop, kid_race, total_pop, deaths_race, late_dx, new_dx, sex_dx, prev, prev_sex, prev_race, v_sup, rec_care)
```

>>>>

Cleaning Part 1

Load necessary libraries
```{r, echo=FALSE, message=FALSE}
library(knitr)
library(readxl)
library(lubridate)
library(stringr)
library(tidyverse)
library(here)
```

Load the files to be cleaned.
```{r}
deaths <- read_xlsx(here("data/raw_data/AIDSVu_State_Death_2020.xlsx"), skip = 3)
us_race_pop <- read_xlsx(here("data/raw_data/DECENNIALPL2020.P2-2023-02-24T124235.xlsx"), sheet = 2)
us_sex_pop <-read_csv(here("data/raw_data/raw_data.csv"), skip = 2)
kid_pop <- read_xlsx(here("data/raw_data/Child population by age group.xlsx"))
kid_race <- read_xlsx(here("data/raw_data/Child population by race and ethnicity.xlsx"))

@@

Let's start by separating `deaths` by demographic characteristic . While this data provides information on demographic characteristics such as sex, age group, and race/ethnicity, it does not provide data on all of these terms together. We'll separating by sex and race/ethnicity. This analysis will be limited to Non-Hispanic Black, Non-Hispanic White, Hispanic persons, and persons of Multiple racial backgrounds.
```{r}
deaths <- deaths %>% 
  filter(!grepl("DC", `State Abbreviation`)) %>%
  filter(!grepl("PR", `State Abbreviation`))

@@

summary(deaths_race)
```
Some rates are -9 due to missingness. Let's convert these to NA
```{r}
deaths_race <- deaths_race %>% 
  mutate(across(ends_with("Death Rate"), ~ case_when( .x < 0 ~ NA, TRUE ~ .x)))
```

Next, let's clean the dataframe for New diagnoses. We'll select for variables of interest and subset.
```{r}
new_dx <- new_dx %>% 
  filter(!grepl("DC", `State Abbreviation`)) %>%
  filter(!grepl("PR", `State Abbreviation`)) %>% 
  select(1:3, contains(c("male", "Black", "White", "Hispanic", "Multiple")), -contains(c("and", "Stability")))

@@

Moving on to the census data. I'll start by re-formatting the data from wide format for States to long, and format race/ethnicity into wide.
```{r}
glimpse(us_race_pop)

@@

I'll select only the columns of interest, and rename some of them so they are a bit easier to work with.
```{r}
race_pop <- race_pop %>% 
  select(c(Location, starts_with(c("Total:", "Black or African American alone", "Hispanic", "White alone", "Population of two or more races:")))) %>% 
  rename(c(Hispanic = `Hispanic or Latino`,
         White = `White alone`,
         Black = `Black or African American alone`,
         Multiple = `Population of two or more races:`))
```

@@

I also need to estimate the population older than 12. Since I cannot find a source explicitly population older than 12, I will estimate population younger than 13 this using data from Kids Count. 
```{r}
kid_pop <- kid_pop %>% 
  filter(TimeFrame == 2020) %>% 
  filter(`Age group` %in% c("0 to 4", "5 to 11", "12 to 14")) %>% 
  mutate(Data = as.numeric(Data))

@@

Then, I'll calculate the counts of males and females in each location using percentage estimates from the KFF and removing the population under 13 years old.
```{r}
new_pop <- list(total_pop, pop_less13, us_sex_pop)

new_pop <- reduce(new_pop, inner_join, by = join_by(Location))

@@

Merging dataframes. 
```{r}
sex_df <- full_join(deaths_sex, df_pop, by = join_by(State == Location))

sex_df <- full_join(sex_df, sex_dx)
```
We need to repeat the steps above, but with race and ethnicity. We'll load the file containing information on the racial/ethnic make up of the population under 18. This contains a lot of data, so we'll be sure to only select the data we need. 
```{r}
glimpse(kid_race)

@@

Let's estimate the number of children under 13 by race. We'll use mutate and across to provide these calculations for each columns for race/ ethnicity. Then, we'll merge and subtract the population under 13 years old from the analysis.
```{r}
kid_race <- kid_race %>%
  mutate(across(2:5, ~ round( . * Less13)))

@@

Merging datasets. 
```{r}
race_df <- list(deaths_race, late_dx, new_dx)

@@

Cleaning the file `Received Care`. I do some cleaning and merge appropriate dataframes, and use some base R to add a couple columns to `sex_df`.
```{r}
rec_care <- rec_care %>%
  filter(!grepl("DC", `State Abbreviation`)) %>%
  filter(!grepl("PR", `State Abbreviation`)) %>% 
  select(1:3, contains("Percent of")) %>% 
  select(1:3, contains(c("Hispanic", "Black", "White", "Multiple","male")))

sex_df$Male_rec_care_percent <- rec_care$`Percent of Males Living with HIV who Received HIV Care`
sex_df$Female_rec_care_percent <- rec_care$`Percent of Females Living with HIV who Received HIV Care`

@@

race_df <- inner_join(race_df, rec_care)
```
Repeat for file on `Viral Suppression`
```{r}
v_sup <- v_sup %>%
  filter(!grepl("DC", `State Abbreviation`)) %>%
  filter(!grepl("PR", `State Abbreviation`)) %>% 
  select(1:3, contains("Percent of")) %>% 
  select(1:3, contains(c("Hispanic", "Black", "White", "Multiple","male")))

@@

Cleaning Prevalence file
```{r}
prev <- prev %>% 
  filter(!grepl("DC", `State Abbreviation`)) %>%
  filter(!grepl("PR", `State Abbreviation`))

@@

Some quick cleaning of the social determinant sof health file, and merging this with our two results dataframes `sex_df` and `race_df`.
```{r}
sdoh <- sdoh %>% 
  filter(!grepl("DC", `State Abbreviation`)) %>% 
   filter(!grepl("PR", `State Abbreviation`))

sex_df <- full_join(sex_df, sdoh)

race_df <- full_join(race_df,sdoh)
```

We have some more cleaning to, but let's save these files for now. 
```{r}
saveRDS(sex_df, here("data/processed_data/sex_df.rds"))
saveRDS(race_df, here("data/processed_data/race_df.rds"))
```

```{r}
rm(deaths, df_pop, kid_pop, pop_sex, pop_less13, us_sex_pop, us_race_pop, new_pop, deaths_sex, race_pop, kid_race, total_pop, deaths_race, late_dx, new_dx, sex_dx, prev, prev_sex, prev_race, v_sup, rec_care, sdoh)
```

### 2023-02-27 code/processing_code/2_cleaning_sex_df.Rmd
eda
comments

```{r}
sex_df <- readRDS(here("data/processed_data/sex_df.rds"))
```
Let's find the HIV death rate by sex
```{r}
sex_df <- sex_df %>% 
  mutate(Male_Rate = round(`Male Death Cases`/pop_male*1e5),
         Female_Rate = round(`Female Death Cases`/pop_female*1e5)) %>% 
  select(!`Year`)
```

Next, let's find the expected number of HIV deaths. First, we'll calculate lambda.
```{r}
lambda_male <- sex_df %>% drop_na(`Male Death Cases`) %>% 
  drop_na(pop_male) %>% 
  summarize(lambda_male = sum(`Male Death Cases`)/sum(pop_male)) %>%
  as.numeric()

@@

```{r}
sex_df <- sex_df %>% 
  mutate(ratio_OE_male = `Male Death Cases` / male_exp_deaths,
         ratio_OE_female = `Female Death Cases` / female_exp_deaths)
```

```{r}
female_df <- sex_df %>% select(1:3, contains("female")) 
male_df <- sex_df %>% select(1:3, !contains("female"))

@@

```

```{r}
saveRDS(sex_df, here("data/processed_data/sex_df_cleaned.rds"))
```

>>>>

```{r}
sex_df <- readRDS(here("data/processed_data/sex_df.rds"))
```

Let's find the HIV death rate by sex. We use round to get x to the closest whole number.
```{r}
sex_df <- sex_df %>% 
  mutate(Male_Rate = round(`Male Death Cases`/pop_male*1e5),
         Female_Rate = round(`Female Death Cases`/pop_female*1e5)) %>% 
  select(!`Year`)
```

Next, let's find the expected number of HIV deaths. First, we'll calculate lambda, which we will then multiply by population to get our expected number of deaths.
```{r}
lambda_male <- sex_df %>%
  drop_na(`Male Death Cases`) %>% 
  drop_na(pop_male) %>% 
  summarize(lambda_male = sum(`Male Death Cases`)/sum(pop_male)) %>%
  as.numeric()

@@

Creating variable for ratio of observed deaths over expected deaths.
```{r}
sex_df <- sex_df %>% 
  mutate(ratio_OE_male = `Male Death Cases` / male_exp_deaths,
         ratio_OE_female = `Female Death Cases` / female_exp_deaths)
```

Let's tidy this all up. We'll separate the dataframe briefly by sex, then re-join!
```{r}
female_df <- sex_df %>% select(1:3, contains("female"), 23:35) 
male_df <- sex_df %>% select(1:3, !contains("female"))

@@

```
Save the file. Next, we'll clean up the file for race/ethnicity.
```{r}
saveRDS(sex_df, here("data/processed_data/sex_df_cleaned.rds"))
```
